% !TeX spellcheck = en_US

\section{Conclusions and Future Work} \label{sec:concl}

This final section synthesizes the main scientific and methodological conclusions derived from the development, optimization, and validation of the \textit{AstroPT MM} architecture. It also outlines the prospects and future lines of research established by this thesis in the field of foundation models applied to astrophysics.

\subsection{Scientific and Methodological Conclusions} \label{sec:concl:sci}

The core achievement of this thesis is the successful design and optimization of \textit{AstroPT MM Autoregressive Opt}, a bimodal foundation model (yet natively prepared for scaling to additional modalities) of $\sim 100.75$ million parameters. This model integrates, for the first time under a self-supervised autoregressive framework, spatial photometric information from \textit{Euclid} and fine spectroscopic information from \textit{\gls{DESI}}. Pretraining was completed on a curated catalog of $137\,661$ unique galaxies, unifying optical and near-infrared images from \textit{Euclid} (MER DR1) with optical spectra from \textit{\gls{DESI}} (\textit{\gls{DESI}} DR1). Based on the systematic evaluation and quantitative and qualitative astrophysical analysis presented in \cref{sec:res}, the following fundamental conclusions are drawn:

\begin{enumerate}
	\item \textbf{Multimodal Synergy and Breaking Observational Degeneracies:} 
	Evaluation via \textit{downstream tasks} demonstrates that the joint latent representation space consistently outperforms individual unimodal modalities. The model achieves an outstanding coefficient of determination in estimating first-order physical parameters, such as total stellar mass ($\log M_*$, $R^2 = 0.9339$) and star formation rate (\textit{log \gls{SFR}}, $R^2 = 0.7487$). 
	
	Physically, this systematic increase in joint-space accuracy confirms that unsupervised bimodal latent fusion breaks degeneracies intrinsic to each instrumental domain. It simultaneously resolves the reddening and age-metallicity degeneracy limiting broadband imaging from \textit{Euclid}, and compensates for the aperture bias of the \textit{\gls{DESI}} fiber ($1.5''$ central diameter), which underestimates the diffuse star formation occurring in the outer disk and spiral arms of galaxies.
	
	\item \textbf{Competitiveness of Continuous Autoregression:} 
	The continuous autoregressive pretraining of \textit{AstroPT MM Autoregressive Opt} provides a robust and competitive framework compared to traditional contrastive alignment schemes. In direct comparison with \textit{AstroCLIP} \autocite{ar:astroclip}, our model matches or improves upon its performance in the joint modality for the vast majority of physical and geometric tasks. 
	
	Crucially, in the pure visual channel, \textit{AstroPT MM} outperforms \textit{AstroCLIP} on key physical targets such as stellar mass ($R^2 = 0.8427$ versus $0.7526$) and \textit{\gls{SFR}} ($R^2 = 0.6398$ versus $0.6037$). This demonstrates that autoregressive learning with token mixing compels the encoder to model high-fidelity morphological features instead of relying on low-frequency global alignments.
	
	\item \textbf{Outperforming the Unimodal Supervised Limit:} 
	One of the critical findings of this project is that the joint-space embeddings of \textit{AstroPT MM Autoregressive Opt} systematically outperform task-specific unimodal supervised reference models (such as the convolutional \textit{ResNet18} network for images and the Conv1D+Transformer model for spectra). 
	
	Although the latter were trained with direct optimization on physical labels for a single task, the bimodal pretraining of \textit{AstroPT MM Autoregressive Opt} instills a density of physical knowledge and cross-modal interconnection that more than compensates for the lack of direct supervision. This validates the foundation model hypothesis in astrophysics: multimodal self-supervised learning yields richer and more generalist latent representations than traditional supervised pretraining.
	
	\item \textbf{Limitations of Hybrid Contrastive Alignment:} 
	The architectural variant \textit{Hybrid AstroPT MM}, which introduces InfoNCE contrastive losses at the midpoint of the transformer (layer 6) using a dual-phase attention mask, reported significantly lower performance in estimating physical variables ($R^2 = 0.8675$ for joint-modality stellar mass and $0.6447$ for \textit{\gls{SFR}}). 
	
	This experimental behavior shows that imposing a global vector similarity bottleneck at the midpoint of the transformer's flow severely restricts the non-linear expressiveness that the bimodal autoregressive model freely exploits from the early layers. Contrastive alignment degrades the detailed physical representation of individual physical flux patches, limiting its ability to model complex transitions on the multimodal latent manifold.
	
	\item \textbf{Self-Attention Dynamics and the $F(\text{H}_\beta)$ Anomaly:} 
	The performance drop experienced by the nebular recombination line flux $\text{H}_\beta$ in the joint space ($R^2 = 0.7374$ compared to the excellent $R^2 = 0.9594$ of the isolated spectroscopic branch) highlights an important limitation in the bimodal optimization dynamics of causal self-attention. During token mixing, the higher \textit{\gls{SNR}} and amplitude of the $\text{H}_\alpha$ line in the red optical range tend to dominate the attention gradients, overshadowing lower-amplitude spectral features in the blue range, such as the nebular hydrogen doublet of $\text{H}_\beta$. This indicates the need to balance loss functions to prevent the model from ignoring fine-grained physical details.
	\item \textbf{Absence of Modal Transfer:} As noted in \cref{sec:res:cross}, knowledge transfer between modalities during training is deficient. While some appreciable transfer exists when comparing the \textit{Autoregressive Opt} model against \textit{SSL Unimodal I} or \textit{Supervised I} (or their spectral counterparts), it is still far from achieving cross-modal transfer that matches performance in tasks where they are not the dominant modality. Improving this deficit will be fundamental, especially to obtain spectroscopic information from images acquired with next-generation surveys.
\end{enumerate}

In conclusion, this work consolidates continuous autoregressive pretraining of bimodal patches as a viable and highly efficient paradigm for constructing unified representation models of the local universe.

\subsection{Future Work and Scaling Perspectives} \label{sec:concl:future}

Future research lines and experimental extensions derived from this work are structured along the following technological and astrophysical axes:

\begin{itemize}
	\item \textbf{Value-Added Catalogs via Latent Transfer:} 
	The primary immediate observational application of \textit{AstroPT MM Autoregressive Opt} is the high-precision inference of physical parameters for galaxies lacking spectroscopic data. By using the visual embeddings from the image encoder (which absorbed spectral knowledge during bimodal pretraining), it is possible to construct accurate estimators for photometric redshifts ($z_{\text{phot}}$), stellar masses, and \textit{\gls{SFR}} from pure optical/near-infrared images, providing a massive synthetic catalog at a scale impossible to achieve using conventional physical instrumentation.
	
	\item \textbf{Unsupervised Scientific Applications in Observational Astronomy:} 
	As documented in \cref{sec:ap:sim}, the richness of the bimodal space opens the door to direct unsupervised exploitation. The $k$-NN similarity search engine implemented on the joint latent space enables ultra-fast searches to identify and catalog rare astrophysical sources of high scientific relevance, such as \textit{dual \glspl{AGN}}.
	
	\item \textbf{Generative Models and Latent Diffusion in Embedding Spaces:} 
	One of the most innovative proposals, experimentally introduced in \cref{sec:ap:diff}, is the development of conditioned latent diffusion models within the latent space of \textit{AstroPT MM Autoregressive Opt}. By training a diffusive model to transition between visual morphology embeddings and the continuous flux spectrum, the direct synthesis of optical spectra conditioned on the visual morphological features of \textit{Euclid} will be enabled. This tool will act as a ``virtual spectrograph'', predicting emission lines and resolved metallicities from wide-field photometry.
	
	\item \textbf{Data Scaling and Modeling in the Chinchilla Limit:} 
	Since the current data volume places pretraining in a suboptimal token regime relative to the model parameters, future work contemplates incorporating the complete overlap with future massive surveys. Currently, the model is prepared to support new data modalities from ground-based surveys, such as those carried out by the \textit{\gls{HSC}} project or the \textit{\gls{SDSS}} survey. Incorporating wide-field survey data from the \textit{Vera C. Rubin Observatory} (\textit{\gls{LSST}}) and the future \textit{Nancy Grace Roman Space Telescope} will provide the volume of objects necessary to reach the optimal convergence zone and classical scaling laws. It is expected that a larger data volume and a richer pretraining set will enable the model to continue improving physical parameter predictions.
	
	\item \textbf{\gls{MoE} Architectures:} 
	To scale the bimodal foundation model beyond 100 million parameters without drastically increasing the computational cost on \textit{TeideHPC}, we propose restructuring the network using a state-of-the-art architecture such as \textit{\gls{MoE}}. In this scheme, specific transformer layers will act as experts specialized in either spatial morphology or fine-grained spectral kinematics, dynamically routing visual and spectral tokens to optimize the physical representation of the joint space.
	
	\item \textbf{Integration with LLMs and Recurrent Architectures:} 
	A future development path with great potential is the alignment and coupling of the physical latent space of \textit{AstroPT MM} with the semantic representation space of science-oriented multimodal \textit{\glspl{LLM}}. Projecting our joint embeddings as input tokens into an \textit{\gls{LLM}} will facilitate interactive physical reasoning and the automatic generation of scientific hypotheses from direct observations. 

	Additionally, the exploration of architectures based on \textit{Looped Transformers} or recurrent \textit{\glspl{LLM}} is contemplated. These architectures, which execute a variable number of inference steps by sharing weights in iterative loops over the token sequence, enable the emulation of adaptive network depth and the execution of dynamic iterative computations (such as galaxy evolution or radiative transfer simulations in the interstellar medium) directly within the latent manifold, transcending the static paradigm of classical autoregression.
	
	\item \textbf{Exploration of Discrete Tokenizers in Large Data Volume Regimes:} 
	Although the initial discretization experiment using the pretrained encoders of \textit{AION-1} proved unfeasible due to quantization noise and instrumental domain mismatch, this approach retains significant potential value for future work with larger-scale samples. 

	With the future incorporation of massive datasets from complete surveys (such as the \textit{Euclid} DR1 releases and the commissioning of the \textit{\gls{LSST}}), we plan to develop and pretrain native bimodal discrete tokenizers from scratch (using techniques such as \textit{\gls{FSQ}} or \textit{\gls{LFQ}}). A critical data volume of several million galaxies will allow vector quantizers to structure dense and robust codebooks, capable of retaining subtle surface brightness gradients and fine spectral indicators without suffering information loss from quantization collapse, effectively enabling the treatment of the observable universe as a unified discrete language.
	
	\item \textbf{Unified Pretraining Strategy on Unpaired Data:} 
	One of the greatest physical bottlenecks for scaling \textit{AstroPT MM} is the restriction of training exclusively on samples with direct overlap (the positional matching catalog of $137\,661$ objects). To bypass this data volume limit, we propose a modification to the data loading pipeline and self-attention processing to allow the model to ingest unpaired data on a massive scale. 

	Under this semi-paired pretraining scheme, the architecture will simultaneously assimilate millions of pure \textit{Euclid} images and isolated \textit{\gls{DESI}} spectra, applying dynamic masks on the transformer to deactivate bimodal self-attention when one modality is absent and optimizing the branches in a unimodal autoregressive manner. During the same pretraining cycle, the optimizer will interleave fully paired data batches to refine, align, and unite the joint latent space using cross-attention and contrastive losses. This strategy will make it possible to exploit the entirety of the data available in each survey independently, maximizing the learning of the foundation model within a unified pipeline.
\end{itemize}