% !TeX spellcheck = en_US

\section*{Acknowledgments}

With the conclusion of this project, and as it coincides with my 25th birthday, I would like to express my gratitude:

\textbf{To my parents}, because without them, nothing I have achieved in life would have been possible. For supporting me at all times since I told them at 8 years old that I wanted to be an astrophysicist, and for providing me with every resource within their reach to achieve it. For nurturing in me since childhood the curiosity that today forms the core of my being, and for being the hands that started my internal engine so that I could walk my own path independently. I will never be able to repay them for everything they have given me. I only hope to be as good a parent as they have been to me.

\textbf{To my partner}, for being a fundamental pillar and creating a daily closeness despite the distance. Because on difficult days, when nothing seems to work, she always brings a smile to my face. For giving me that extra boost when energy is running low, and for multiplying my joy by far when we are together. For embarking with me on this adventure of uncertainty and doing her part so that I can fulfill my dream.

\textbf{To my grandparents}, because although they have never fully understood what exactly I do, that has never prevented them from sharing the joy and excitement of my academic achievements. For giving me the best summers of my childhood in the village alongside them, and for providing the affection that every grandchild receives from good grandparents.

\textbf{To my friends}, to all those I have met at some point along the journey and especially to those who remain in my life. Because despite the distance and not being able to dedicate all the time I would like to them, I know I can still count on them just like on day one. I am truly fortunate for the friendships I have; I would not change them for anything.

\textbf{To the teachers} of the different stages of my academic life. For having guided me in the academic world and pointing out the best paths to travel toward this goal. I will never forget their contribution to the achievement of my past and future milestones.

\textbf{To life}, because although the four knee surgeries were a major obstacle, it did not let me lose the enthusiasm to reach where I am today. For giving me opportunities that few people can access, and which over the years I feel increasingly fortunate to have experienced. For putting the right people in my path and for providing the conditions to ensure that I have been, and am, very happy in this quarter of a century of my existence.

\textbf{To the Víctor of the past}, for his resilience in the most complicated moments. For learning to walk again after each surgery and keeping alive the flame of the goal set more than fifteen years ago. For being so multifaceted that, although having so many interests can sometimes be a bit chaotic, it has given me a backpack full of varied experiences. I hope, \textbf{Víctor of the future}, that this curiosity and this drive to pursue motivating and varied projects remain over time. And I hope, Víctor of the future, that you manage to finish the remaining third of the path we have left to travel. But remember: as long as you fight for it, even if you do not succeed, if you gave everything you had, you should be proud. \textit{The final step that crosses the finish line is just as important as all those that bring you closer to it.}



\vspace{2cm}
\noindent
\begin{minipage}[t]{\textwidth}
	\raggedleft
	In San Cristóbal de La Laguna, July 9, 2026.
\end{minipage}